---
author: 
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(lubridate)
```

## Assignment 1

#### Due Friday, September 17, 2021, at 11:59 PM

### Preliminaries

- Verify this file (`hw01.Rmd`) is in the directory `COURSE/hw/hw01/`.
- Set the working directory to the same.

## Problems

### 1

Use R to sum up the numbers from 1 to 100 by:

- creating a vector with these numbers
- using the base R function `sum()` to calculate their sum.

```{r}
sum(seq(1, 100))
```

The sum of the integers from 1 to 100 is:

5050

### 2

The following R chunk will set a random seed and then create a vector of random numbers.
By setting the random seed, each person's vector will be identical as all will use the same stream of computer-generated random numbers.

```{r}
set.seed(20210910)
v2 = runif(10)
v2
```

Create a vector named `v2_a` from `v2` by dividing each value by the sum of these ten values. *Note: this re-weighting results in a new vector where the values are proportional to the originals, but sum to one.*

```{r}
v2_a = v2/sum(v2)
v2_a
max(v2_a)
sd(v2_a)
```

#### 2A

What is the first element of `v2_a`?

0.1087

#### 2B

What is the maximum value of `v2_a`?

0.1700408

#### 2C

What is the standard deviation of `v2_a`? Use the function `sd()`.
(The standard deviation is a numerical summary of a sample of numbers that measures the size of the spread around their mean.)

0.0632675220345583

#### 2D

Verify that the sum of `v2_a` is, in fact, equal to 1.

```{r}
sum(v2_a)
```

### 3

This code will generate a sample of 100 randomly generated standard normal random variables saved as a vector `x`.

```{r}
x = rnorm(100)
```

Use R to calculate the the sum of the squares of these values.

$$
\sum_{i=1}^{100} x_i^2
$$ 

(*The previous code is an example of using LaTeX within R Markdown to create a formatted mathematical expression in the output HTML document.*)

```{r}
sum(x^2)
```

The sum of squared values is:

111.4508

### 4

This block of code reads in the Lake Mendota winter freeze data.

```{r}
mendota = read_csv("../../data/lake-mendota-winters-2021.csv")
```

### 4A

Create a vector named `y` which is the duration of total days that Lake Mendota is closed by ice for each winter in the data set.

Calculate the mean of these values.

```{r}
thaw = mendota$last_thaw
freeze = mendota$first_freeze
y = as.numeric(thaw-freeze)
mean(y)
```

What is the mean?

102.6687

### 4B

Create a vector named `r` by taking `y` and subtracting from each value the mean you computed in the previous part.

```{r}
r = y-mean(y)
mean(r)
sd(r)
sd(y)
```

What is the mean of `r`?
What are the standard deviations of `r` and `y`?

mean r is -5.640704e-15
standard deviations of r and y is 18.83401

### 4C

Create a vector named `z` by taking `r` and dividing each element by the standard deviation of `r` (computed with `sd()`).

```{r}
z = r/sd(r)
mean(z)
median(z)
sd(z)
```

What are the mean, median, and standard deviation of `z`?

mean is -2.851517e-16
median is 0.07068731
sd is 1





