---
title: "Assignment 08"
output: html_document
editor_options: 
  chunk_output_type: inline
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(tidyverse)
source("../../scripts/viridis.R")
source("../../scripts/ggprob.R")
```

\newcommand{\E}{\mathsf{E}}
\newcommand{\Var}{\mathsf{Var}}
\renewcommand{\prob}{\mathsf{P}}

## Homework 8

#### Due Friday, November 19, 11:59 PM CT

Problems 1-3 are based on the random variable $X \sim \text{Binomial}(60, 0.4)$.

### Problem 1

### A

Calculate and report the mean and standard deviation of the distribution.

```{r}
mean = 60*0.4
mean

sd = sqrt(60*0.4*(1-0.4))
sd
```

### B

Calculate and report $\prob(X = 25)$

```{r}
pbinom(25, 60, 0.4)
```

### C

Calculate and report $\prob(X >= 25)$

```{r}
1-pbinom(25, 60, .4)
```

### D

Create a graph which displays the binomial distribution with bars for $x \ge 25$ colored in a different color from the rest to help visualize your answer to **C**.

```{r}
gbinom(60, 0.4, a = 0, b = 40) +
  geom_binom_density(60, 0.4, a = 1, b = 24, color = "gray") +
  geom_binom_density(60, 0.4, a = 25, b = 60, color = "red") +
  theme_minimal()
```

### Problem 2

What is the probability that $X$ is within one, two, and three standard deviations of the mean? Round each probability to four decimal places.

one sd: 0.6826
two sd: 0.9544
three sd: 0.9972


### Problem 3

Draw a graph of this binomial distribution with vertical lines with different line types at the end points of the intervals 1, 2, and 3 sigma from the mean.

- solid for $\mu \pm \sigma$
- dashed for $\mu \pm 2\sigma$
- dotted for $\mu \pm 3\sigma$

```{r}
mu = 60*0.4
sd = sqrt(60*0.4*(1-0.4))

gbinom(60, 0.4, scale = TRUE) +
  geom_vline(xintercept = mu, color = "red", alpha = 0.5,
             linetype = "dashed") +
  geom_vline(xintercept = mu + c(-1,1)*sd,
             color = "red", linetype = "dashed") +
  geom_vline(xintercept = mu + c(-2,2)*sd,
             color = "red", linetype = "dashed") +
  geom_vline(xintercept = mu + c(-3,3)*sd,
             color = "red", linetype = "dotted") +
  theme_minimal()
```



### Problem 4

### A

Create a tibble where each row corresponds to a binomial distribution with the following columns:

- `n` for $n = 1000,1001, \ldots, 10,000$
- `p` equal to 0.9 for each row
- `mu` equal to the mean
- `sigma` equal to the standard deviation
- `q.975` equal to the 0.975 quantile of the distribution
- `z = (q.975 - mu)/sigma`, the standardized z-score of the quantile
- `prob` equal to $\prob(X \ge q.975)$, a right-tail probability for the distribution in each row

Display all columns and the rows of this table for which $n$ is a multiple of 1000.

```{r}
p4a = tibble(n = seq(1000, 10000), p = 0.9, mu = n*p, sigma = sqrt(n*p*(1-p)),
       q.975 = qbinom(0.975, n, p), z = (q.975-mu)/sigma, 
       prob = 1-pbinom(q.975, n, p))

p4a %>% filter(n %% 1000 == 0)
```

### B

Use a line plot to show the relationship between $n$ on the x axis and $z$ on the y axis.
Add a smooth curve to highlight the trend.
Describe the pattern you see.

```{r}
ggplot(p4a, aes(x = n, y = z)) +
  geom_point() +
  geom_smooth()
```

### C

What is the average of the values of $z$ for $n > 9000$?

```{r}
p4c = p4a %>% 
  filter(n > 9000) %>% 
  summarize(average = mean(z))
p4c
```

### D

Plot the right tail probability `prob` on y axis versus `n` on the x axis.

```{r}
ggplot(p4a, aes(x = prob, y = n)) +
  geom_point() +
  geom_line()
```

### E

What is the average of the values of `prob` for $n > 9000$?

```{r}
p4e = p4a %>% 
  filter(n > 9000) %>% 
  summarize(average = mean(prob))
p4e
```


### Problem 5

### A

What is the 0.975 quantile of the standard normal distribution?

```{r}
qnorm(0.975, 0, 1)
```

### B

What is the area to the right of this quantile?

```{r}
1-pnorm(qnorm(0.975, 0, 1), 0, 1)
```


### Problem 6

For the $\text{N}(500,100)$ distribution, find values $a$ and $b$ such that $a < b$, the area between $a$ and $b$ under the normal density is equal to 0.6,
and $a$ and $b$ are equally distant from $\mu = 500$.

```{r}
qnorm(0.2, 500, 100)
qnorm(0.2, 500, 100, lower.tail = F)
```


### Problem 7

Graph the $\text{N}(500,100)$ distribution and shade the area between $a$ and $b$ from the previous problem.

```{r}
a = qnorm(0.2, 500, 100)
b = qnorm(0.2, 500, 100, lower.tail = F)

gnorm(500, 100) +
  geom_norm_fill(500, 100, a, b)

```





